Towards Artificial Intelligence for Metaphysics: \\ the Case of the Inconsistency in G\"odel's Ontological Argument

### 2016-01-09 Papers/InProgress/IJCAI/paper.tex
Add footnote about personal communication with Dana Scott.
Write section about the inconsistency

\usepackage{amsmath}
\usepackage{xspace}
\usepackage{modallogics}
\usepackage{graphicx,url}
\usepackage{txfonts} % needed for \Diamondblack
\usepackage{float}

@@

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as the ancestor of modern ontological arguments such as G\"odel's. In the millenium between Anselm and G\"odel, many philosophers modified and arguably improved Anselm's argument. Of particular importance to G\"odel was the work of Leibniz \cite{ToDo}. 
G\"odel's manuscript (Figure \ref{GoedelScript}) can be considered a translation of Leibniz's presentation of the argument into modern modal logic. G\"odel shared his manuscript with Scott, who shared a slightly different version with a larger public. Scott's version of the axioms and definitions, formalized in Isabelle, is shown in Figure \ref{ScottS5}. The main difference to G\"odel's version is the addition of a conjunct in the definition of \emph{essence}. G\"odel's different definition of essence can be seen either in his manuscript (Figure \ref{GoedelScript}) or, in more modern notation, in the Isabelle formalization shown in Figure \ref{nconsistencyIsabelleK}. For Scott, an essential property of an individual must be possessed by him/her. For G\"odel, this is not required. This difference has been considered inessential and merely an oversight by many. For instance, Hazen 1998, p.365 \cite[p.365]{Hazen1998} states that ``G\"odel left this clause out in [11], but this appears to have been an oversight--it is included in related manuscripts''. However, the omitted conjunct is in fact crucial. Without it, G\"odel's original axioms are inconsistent. With it, Scott's axioms are consistent.

@@

\section{Intuitive Inconsistency Argument}
\subsection{LEO-II's Inaccessible Inconsistency Proof}
Tell the story (here or earlier). Show parts of LEO proof.  Inaccessible, but relevant knowledge
is contained.
\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./Images/LEO-Proof.png}}
\caption{Primitive substitution in LEO-II generates candidates for the
empty property.} \label{LEO-Proof}
\end{figure*}
LEO-II's resolution proof is human unintuitive. However, it contained
relevant hints to the empty property (cf. Fig.~\ref{LEO-Proof}).

is presented
  for higher-order modal logics KB and K (with possibilist and
  actualist first-order quantifiers)

\subsection{Argument Reconstruction in Isabelle}
Present the argument informal and in Isabelle (see
Fig.~\ref{InconsistencyIsabelleK}). Easy to understand in 
in KB and KT, slightly harder in K.
\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Reconstruction of inconsistency in Isabelle/HOl..} \label{InconsistencyIsabelleK}
\end{figure}

>>>>

\usepackage{amsmath}
\usepackage{xspace}
\usepackage{modallogics}
\usepackage{graphicx,url}
\usepackage{txfonts} % needed for \Diamondblack
\usepackage{color}
\usepackage{xcolor}

@@

\newcommand{\NE}{\mathit{NE}}
\newcommand{\ess}[2]{#1 \mathit{ess} #2}
\newcommand{\nec}{\Box}
\newcommand{\pos}{\Dia}

@@

St. Anselm's ontological argument \cite{Proslogion} can be regarded as the ancestor of modern ontological arguments such as G\"odel's. In the millenium between Anselm and G\"odel, many philosophers modified and arguably improved Anselm's argument. Of particular importance to G\"odel was the work of Leibniz \cite{ToDo}. 
G\"odel's manuscript (Figure \ref{GoedelScript}) can be considered a translation of Leibniz's presentation of the argument into modern modal logic. G\"odel shared his manuscript with Scott, who shared a slightly different version with a larger public. Scott's version of the axioms and definitions, formalized in Isabelle, is shown in Figure \ref{ScottS5}. The main difference to G\"odel's version is the addition of a conjunct in the definition of \emph{essence}. G\"odel's different definition of essence can be seen either in his manuscript (Figure \ref{GoedelScript}) or, in more modern notation, in the Isabelle formalization shown in Figure \ref{nconsistencyIsabelleK}. For Scott, an essential property of an individual must be possessed by him/her. For G\"odel, this is not required. This difference has been considered inessential and merely an oversight by many. For instance, Hazen 1998, p.365 \cite[p.365]{Hazen1998} states that ``G\"odel left this clause out in [11], but this appears to have been an oversight--it is included in related manuscripts''. However, the omitted conjunct is in fact crucial. Without it, G\"odel's original axioms are inconsistent. With it, Scott's axioms are consistent\footnote{In personal communication, Dana Scott confirmed that he was unaware G\"odel's axioms were inconsistent.}.

@@

\section{Intuitive Inconsistency Argument}

In the typical workflow of trying to prove a conjecture with a theorem prover, it is customary to check the consistency of the axioms first. For if the axioms are inconsistent, anything (including the conjecture) would be trivially derivable in classical logic (\emph{ex falso quodlibet}). Surprisingly, when this routine check was performed on G\"odel's axioms \cite{ECAI}, the LEO-II prover claimed that the axioms were inconsistent. Unfortunately, the refutation generated by LEO-II was barely human-readable. The text file was 153 lines\footnote{Long lines with an average of 184 characters per line.} long and machine-oriented proof calculus (higher-order resolution \cite{}) and syntax (TPTP THF \cite{}) were used. Part of the file is displayed in Fig.~\ref{LEO-Proof}.

\begin{figure*}
\centerline{\includegraphics[width=\textwidth]{./Images/LEO-Proof.png}}
\caption{Lines 115--120 of LEO-II's Refutation. Primitive Substitutions with Empty Property are Highlighted.} \label{LEO-Proof}
\end{figure*}

Although LEO-II's resolution refutation is not easy to read for humans, it did contain
relevant hints to the importance of the empty property (cf. Fig.~\ref{LEO-Proof}).

\subsection{Informal Refutation}

Based on the hints found in LEO-II's refutation, we conceived the following informal explanation for the inconsistency of G\"odel's axioms:

\begin{enumerate}
\item From G\"odel's definition of essence 
(${\ess{\phi}{x} \biimp {\allq \psi} (\psi(x)
\imp {\nec} \allq y (\phi(y) \imp \psi(y)))}$) it follows that the empty property is an essence of every individual (\textbf{Empty Essence Lemma}): 
$$\allq x\,(\ess{\emptyset}{x})$$.

\item From theorem T1 (positive Properties are possibly
  exemplified: ${\allq \phi} [P(\phi) \imp {\pos}  \exq x
  \phi(x)]$) and axiom A5 (``necessary existence'' is a positive property: $P(\NE)$ ), it follows that $\NE$ is possibly exemplified:
  $$
  \pos \exq x [\NE(x)]
  $$

\item Expanding the definition of ``necessary existence'' ( 
  ${\NE(x) \equiv \allq
  \phi [\ess{\phi}{x} \imp \nec
  \exq y \phi(y)]}$), the following is obtained:
  $$
  \pos \exq x [\allq \varphi [ \ess{\varphi}{x} \imp \nec \exq y [\varphi(y)] ] ]
  $$

\item The sentence above holds for all $\varphi$ and thus, in particular, for the empty property:
$$
\pos \exq x [ \ess{\emptyset}{x} \imp \nec \exq y [\emptyset(y)] ]
$$

\item By the Empty Essence Lemma, the antecedent of the implication above is valid. Therefore, the sentence above entails:
$$
\pos \exq x [ \nec \exq y [\emptyset(y)] ]
$$

\item By definition of $\emptyset$: 
$$
\pos \exq x [ \nec \bot ]
$$

\item As the existential quantifier is binding no variable within its scope, the sentence is equi-valid with:
$$\pos \nec \bot $$

\item To see that the sentence above is contradictory, we may reason semantically, in terms of possible worlds. If $w_0$ is the arbitrary current world, the $\pos$ operator forces the existence of a world $w$ accessible from $w_0$ such that $\nec \bot$ is true in $w$. But $\nec \bot$ can only be true in $w$, if there is no world $w'$ accessible from $w$. In logics with a reflexive or symmetric accessibility relation (e.g. \KB), it is easy to see that there must be a world $w'$ accessible from $w$: either $w'$ itself, in case of a reflexive relation, or $w_0$, in case of a symmetric relation. In fact, even in \K, with no accessibility condition, there must be a world $w'$ accessible from $w$. The reason is that $\pos \nec \bot$ should be \emph{valid} (true in all worlds). Therefore, it is true in $w$ as well, where the existence of an accessible world $w'$ is forced by the $\pos$ operator. As a model for $\pos \nec \bot$ (which is a consequence of G\"odel's axioms) cannot be built, G\"odel's axioms are inconsistent.
\end{enumerate}

Interestingly, the refutation automatically generated by LEO-II uses a symmetric accessibility relation, and thus requires the modal logic KB. The informal refutation described above, on the other hand, requires only the weaker modal logic K.

\subsection{Argument Reconstruction in Isabelle}

To verify the correctness of the informal argument explained above, it was reconstructed in Isabelle/HOL, using Metis to automate the inessential parts. The essential use of the Empty Essence Lemma, on the other hand, is explicitly stated, to ensure that Isabelle is reconstructing the same argument. In fact, without the help of this lemma, Metis is not strong enough to refute G\"odel's axioms. \marginpar{ToDo: check this claim about Metis.}

\begin{figure}
\centerline{\includegraphics[width=\columnwidth]{./Images/InconsistencyIsabelleK.png}}
\caption{Refutation of G\"odel's Axioms in Isabelle/HOl..} \label{InconsistencyIsabelleK}
\end{figure}
